---
title: Lipidomics report
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: false
output_file: lip
params: 
  metadata_file: ../data/metadata.txt
  input_file: ../data/molecular_species_data_SRM_AA_T1D.csv
  meta_sample_name: SampleName
  meta_condition_type: Type
  meta_rep: Rep
  lipid_name_col: X1
  paired: FALSE
  quick_test: TRUE
  output_folder: out
  bar_class_ylabel: value
  control: SRM

---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = FALSE)
library(ggplot2)
library(tidyverse)
options(knitr.duplicate.label = "allow")
source("heatmap_processing.R")
source("stats.R")
source("utils.R")
```

```{r import_data, include=FALSE}
all_data <- switch(tools::file_ext(params$input_file), 
       tsv = ,
       txt = readr::read_tsv(params$input_file),
       csv = readr::read_csv(params$input_file),
       stop("Unknown file extension on data file")
)

meta_info <- process_metadata(metadata_file = params$metadata_file, control = params$control)

exp_conditions <- pull(distinct(meta_info, condition))

```

```{r checks, include = FALSE}
# sanity checks on data needed
assertthat::assert_that(
  params$lipid_name_col %in% colnames(all_data), 
  msg = paste0(
    "params$lipid_name_col ", 
    params$lipid_name_col, 
    " not found in ", 
    colnames(all_data)
  )
)

assertthat::assert_that(
  all(colnames(all_data) %in% c(meta_info$sample_name, params$lipid_name_col)),
  msg = paste0("Unexpected column name found in all_data file")
)
```

```{r process_data, include = FALSE}
all_data <- dplyr::rename(
  all_data, lipid_name := tidyselect::all_of(params$lipid_name_col)
)

all_data <- all_data %>%
  replace(is.na(.), 0)

if(params$paired) {
  meta_type <- meta_info %>%
    dplyr::rename(rep = tidyselect::all_of(params$meta_rep)) %>%
    dplyr::select(sample_name, condition, rep) %>%
    dplyr::mutate(rep = forcats::as_factor(rep))
} else {
  meta_type <- meta_info %>%
    dplyr::select(sample_name, condition)
}

tidy_data <- all_data %>%
  tidyr::pivot_longer(cols = -lipid_name, names_to = "sample_name") %>%
  replace(is.na(.), 0) %>%
  tidyr::separate(lipid_name, sep = " ", into = c("class", NA), remove = FALSE) %>%
  left_join(meta_type) 
```

# Checking skew  

Skewness > 1 suggests positive skew which suggests log-normality.  
One way to check it out is to look at the minimum skewness.value.

```{r}
tidy_data %>%
  group_by(condition, sample_name) %>%
  summarise(skew.value = PerformanceAnalytics::skewness(value)) %>%
  rstatix::get_summary_stats(skew.value) %>%
  DT::datatable(rownames = FALSE, options = list(dom = "t"))
```

All minimum skewness values are well above 1 so we'll log transform the data.

```{r}
tidy_data <- tidy_data %>%
  dplyr::mutate(log2_value = log2(value)) %>%
  na_if("-Inf")
  #dplyr::mutate(log2_value = if_else(log2_value == "-Inf", 0, log2_value)) # we don't want -Inf values
```

Check whether the log transformation has improved the skew values.
```{r}
tidy_data %>%
  group_by(condition, sample_name) %>%
  summarise(log.skew.value = PerformanceAnalytics::skewness(log2_value)) %>%
  rstatix::get_summary_stats(log.skew.value) %>%
  DT::datatable(rownames = FALSE, options = list(dom = "t"))
```
The max skew values are all below 1, so we'll continue with the log transformed data.


```{r process_bar_data, eval = FALSE}
# this needs to be done here, not in the bar_plot Rmd file, so that it is only 
# run once
dodge_width <- 0.8

bar_data_linear_median <- tidy_data %>%
  dplyr::group_by(class, lipid_name, condition) %>%
  dplyr::summarise(median = median(value)) %>%
  ungroup() %>%
  group_by(lipid_name) %>%
  mutate(min_median = min(median)) %>%
  mutate(max_median = max(median)) %>%
  ungroup() %>%
  mutate(abundance = if_else(max_median <= 1, "low", "high")) %>%
  left_join(tidy_data) %>%
  select(class, lipid_name, value, median, condition, abundance) 
  
```


```{r anova_test, eval=FALSE}
## test with 3 species with no missing data ##
tidy_data %>%
  filter(lipid_name =="DG 32:1" |lipid_name == "DG 34:1" | lipid_name == "DG 34:2") %>%
  group_by(lipid_name) %>%
    rstatix::anova_test(log2_value ~ condition)

## it works ##

## now let's remove species with more than 2 NA per condition ##
tidy_data %>%
  filter(lipid_name =="DG 32:1" |lipid_name == "DG 34:1" | lipid_name == "DG 34:2" | lipid_name == "DG 32:0") -> tidy.sample

tidy.sample %>%
  group_by(lipid_name) %>%
  summarise(na_count = sum(is.na(log2_value))) %>%
  filter(na_count <= 1) %>%
  left_join(tidy.sample) -> tidy.sample


#### We need to use anova_test but we need to remove the species with too many the NA first: na> 3? ###
anova_test_res <- tidy_data %>%
    rstatix::anova_test(log2_value ~ condition)
```

```{r}
anova_res <- tidy_data %>%
  group_by(lipid_name) %>%
  summarise(na_count = sum(is.na(log2_value))) %>%
  filter(na_count <= 1) %>%
  ungroup() %>%
  left_join(tidy_data) %>%
  group_by(lipid_name) %>%
  rstatix::anova_test(log2_value ~ condition) %>%
  ungroup()

sig_lipids <- anova_res[anova_res$p < 0.05, ] # tidyverse complains about the anova format so we'll go old school

# dplyr::filter(p < 0.05)
```

```{r}
sig_lipids %>%
  DT::datatable(rownames = FALSE, options = list(pageLength = 10, dom = "tp"))
```


## Summary

```{r, child='summary_plots.Rmd'}
```


## Experimental info

<div class = "row">
<div class = "col-md-6">
```{r}
DT::datatable(meta_type, rownames = FALSE, options = list(pageLength = 15, dom = "tp"))
```
</div>
</div>
<br>
   
### QQ plots of all data  
  
0 values have been removed.  

```{r, out.width=c('50%', '50%'), fig.show='hold', fig.height = 6}
removed_0 <- tidy_data %>%
  filter(value > 0)

qqnorm(removed_0$value, main = "linear/raw")
qqline(removed_0$value, col = "red", lwd = 2)

qqnorm(removed_0$log2_value, main = "log2 transformed")
qqline(removed_0$log2_value, col = "red", lwd = 2)
```
