---
title: Lipidomics report
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: false
output_file: lip
params: 
  metadata_file: ../data/metadata.txt
  input_file: ../data/molecular_species_data_SRM_AA_T1D.csv
  meta_sample_name: SampleName
  meta_condition_type: Type
  meta_rep: Rep
  lipid_name_col: X1
  paired: FALSE
  quick_test: TRUE
  output_folder: out
  bar_class_ylabel: value
  control: SRM

---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = FALSE)
library(ggplot2)
library(tidyverse)
options(knitr.duplicate.label = "allow")
source("heatmap_processing.R")
source("stats.R")
source("utils.R")
```

```{r import_data, include=FALSE}
all_data <- switch(tools::file_ext(params$input_file), 
       tsv = ,
       txt = readr::read_tsv(params$input_file),
       csv = readr::read_csv(params$input_file),
       stop("Unknown file extension on data file")
)

meta_info <- process_metadata(metadata_file = params$metadata_file, control = params$control)

exp_conditions <- pull(distinct(meta_info, condition))

if(! dir.exists(params$output_folder)) dir.create(params$output_folder)

#TODO:
## save pie chart as pdfs
## "other" set as separate pie charts
## pairwise SRM vs AA and SRM vs T1D
## correlation matrix for pairwise comparisons
## ## save data tables as csv in this and the other report
##  # anova for higher level - at class level
##  # effect of length 32:1 - length 32, double bond 1, pairwise.
```

```{r checks, include = FALSE}
# sanity checks on data needed
assertthat::assert_that(
  params$lipid_name_col %in% colnames(all_data), 
  msg = paste0(
    "params$lipid_name_col ", 
    params$lipid_name_col, 
    " not found in ", 
    colnames(all_data)
  )
)

assertthat::assert_that(
  all(colnames(all_data) %in% c(meta_info$sample_name, params$lipid_name_col)),
  msg = paste0("Unexpected column name found in all_data file")
)
```

```{r process_data, include = FALSE}
all_data <- dplyr::rename(
  all_data, lipid_name := tidyselect::all_of(params$lipid_name_col)
)

all_data <- all_data %>%
  replace(is.na(.), 0)

if(params$paired) {
  meta_type <- meta_info %>%
    dplyr::rename(rep = tidyselect::all_of(params$meta_rep)) %>%
    dplyr::select(sample_name, condition, rep) %>%
    dplyr::mutate(rep = forcats::as_factor(rep))
} else {
  meta_type <- meta_info %>%
    dplyr::select(sample_name, condition)
}

tidy_data <- all_data %>%
  tidyr::pivot_longer(cols = -lipid_name, names_to = "sample_name") %>%
  replace(is.na(.), 0) %>%
  tidyr::separate(lipid_name, sep = " ", into = c("class", NA), remove = FALSE) %>%
  left_join(meta_type) 
```

# Checking skew  

Skewness > 1 suggests positive skew which suggests log-normality.  
One way to check it out is to look at the minimum skewness.value.

```{r}
tidy_data %>%
  group_by(condition, sample_name) %>%
  summarise(skew.value = PerformanceAnalytics::skewness(value)) %>%
  rstatix::get_summary_stats(skew.value) %>%
  DT::datatable(rownames = FALSE, options = list(dom = "t"))
```

All minimum skewness values are well above 1 so we'll log transform the data.

```{r}
tidy_data <- tidy_data %>%
  dplyr::mutate(log2_value = log2(value)) %>%
  na_if("-Inf") %>%
  mutate(condition = fct_relevel(condition, "SRM"))
  #dplyr::mutate(log2_value = if_else(log2_value == "-Inf", 0, log2_value)) # we don't want -Inf values
```

Check whether the log transformation has improved the skew values.
```{r}
tidy_data %>%
  group_by(condition, sample_name) %>%
  summarise(log.skew.value = PerformanceAnalytics::skewness(log2_value)) %>%
  rstatix::get_summary_stats(log.skew.value) %>%
  DT::datatable(rownames = FALSE, options = list(dom = "t"))
```
The max skew values are all below 1, so we'll continue with the log transformed data.


```{r process_bar_data, eval = FALSE}
# this needs to be done here, not in the bar_plot Rmd file, so that it is only 
# run once
dodge_width <- 0.8

bar_data_linear_median <- tidy_data %>%
  dplyr::group_by(class, lipid_name, condition) %>%
  dplyr::summarise(median = median(value)) %>%
  ungroup() %>%
  group_by(lipid_name) %>%
  mutate(min_median = min(median)) %>%
  mutate(max_median = max(median)) %>%
  ungroup() %>%
  mutate(abundance = if_else(max_median <= 1, "low", "high")) %>%
  left_join(tidy_data) %>%
  select(class, lipid_name, value, median, condition, abundance) 
  
```


```{r anova_test, eval=FALSE}
## test with 3 species with no missing data ##
tidy_data %>%
  filter(lipid_name =="DG 32:1" |lipid_name == "DG 34:1" | lipid_name == "DG 34:2") %>%
  group_by(lipid_name) %>%
    rstatix::anova_test(log2_value ~ condition)

## it works ##

## now let's remove species with more than 2 NA per condition ##
tidy_data %>%
  filter(lipid_name =="DG 32:1" |lipid_name == "DG 34:1" | lipid_name == "DG 34:2" | lipid_name == "DG 32:0") -> tidy.sample

tidy.sample %>%
  group_by(lipid_name) %>%
  summarise(na_count = sum(is.na(log2_value))) %>%
  filter(na_count <= 1) %>%
  left_join(tidy.sample) -> tidy.sample


#### We need to use anova_test but we need to remove the species with too many the NA first: na> 3? ###
anova_test_res <- tidy_data %>%
    rstatix::anova_test(log2_value ~ condition)
```

# Anova on total class data 

```{r}
classes <- tidy_data %>%
  select(class, lipid_name) %>%
  distinct()

total_by_class <- tidy_data %>%
  group_by(class) %>%
  summarise(na_count = sum(is.na(log2_value))) %>%
  filter(na_count <= 1) %>%
  ungroup() %>%
  left_join(tidy_data) %>%
  group_by(class, condition, sample_name) %>%
  summarise(total = sum(value)) %>%
  ungroup() %>%
  mutate(log2_total = log2(total))

anova_res_class <- total_by_class %>%
  group_by(class) %>%
  rstatix::anova_test(log2_total ~ condition) %>%
  ungroup() %>%
  as_tibble()

sig_by_class <- anova_res_class %>%
  select(class, p) %>%
  right_join(total_by_class) %>%
  filter(p < 0.05)

```

```{r}
outfile <- paste0(params$output_folder, "/all_anova_class_results.csv")
readr::write_csv(anova_res_class, file = outfile)

anova_res_class %>%
  DT::datatable(rownames = FALSE, options = list(pageLength = 10, dom = "tp"))
```

  
## Anova at individual lipid level

```{r}
by_lipid_name <- tidy_data %>%
  group_by(lipid_name) %>%
  summarise(na_count = sum(is.na(log2_value))) %>%
  filter(na_count <= 1) %>%
  ungroup() %>%
  left_join(tidy_data)

anova_res <- by_lipid_name %>%
  group_by(lipid_name) %>%
  rstatix::anova_test(log2_value ~ condition) %>%
  ungroup() %>%
  as_tibble()

#sig_lipids <- anova_res[anova_res$p < 0.05, ] # tidyverse complains about the anova format so we'll go old school

sig_lipids <- anova_res %>%
  select(lipid_name, p) %>%
  right_join(by_lipid_name) %>%
  filter(p < 0.05)


# dplyr::filter(p < 0.05)
```

We perform an anova on all the lipids to see if there is a difference between any of the conditions.  
  
```{r}
anova_res %>%
  DT::datatable(rownames = FALSE, options = list(pageLength = 10, dom = "tp"))
```

```{r}
outfile <- paste0(params$output_folder, "/all_anova_results.csv")
readr::write_csv(anova_res, file = outfile)
```

## Post hoc tests  

We can run further tests on the lipids identified as changing significantly from the anova.

### Tukey on total class data

```{r}
tukey_by_class <- sig_by_class %>%
  group_by(class) %>%
  rstatix::tukey_hsd(log2_total ~ condition) %>%
  filter(group1 != "AA") %>%
  ungroup() %>%
  as_tibble()

tukey_by_class %>%
  ggplot(aes(class, estimate, ymin=conf.low, ymax=conf.high, fill=group2))+
  geom_errorbar(colour="black", size=0.5, position=position_dodge(.9))+
  geom_col(position = position_dodge(.9), colour="black")+
  scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, face = "bold"))

```

### Tukey on individual lipid data

```{r, fig.width = 10, fig.height=8}
tukey_by_lipid <- sig_lipids %>%
  group_by(lipid_name) %>%
  rstatix::tukey_hsd(log2_value ~ condition) %>%
  filter(!(group1 == "AA" & group2 == "T1D")) %>%
  ungroup() %>%
  as_tibble()

classes <- select(sig_lipids, lipid_name, class)

tukey_by_lipid %>%
  left_join(classes) %>%
  left_join(superclasses) %>%
  ggplot(aes(lipid_name, estimate, ymin=conf.low, ymax=conf.high, fill=group2))+
  geom_errorbar(colour="black", size=0.5, position=position_dodge(.9))+
   geom_col(position = position_dodge(.9), colour="black")+
  scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, face = "bold")) +
  #facet_wrap(~ superclass, scales = "free") +
  facet_wrap(~ class, scales = "free")

# tukey_by_lipid %>%
#   left_join(classes) %>%
#   filter(p.adj < 0.05) %>%
#   ggplot(aes(lipid_name, estimate, ymin=conf.low, ymax=conf.high, fill=group2))+
#   geom_errorbar(colour="black", size=0.5, position=position_dodge(.9))+
#    geom_col(position = position_dodge(.9), colour="black")+
#   scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 90, face = "bold")) +
#   facet_wrap(~ class, scales = "free")

```


## Summary

```{r, child='summary_plots.Rmd'}
```

### Chain length and double bonds

The total amount of lipid varies quite significantly between conditions, so if 
we're comparing amounts between different conditions we should probably normalise
to the largest amount.

```{r}
condition_totals <- tidy_data %>%
  group_by(condition) %>%
  summarise(total = sum(value)) %>%
  mutate(scale_factor = max(total)/total)

DT::datatable(condition_totals, rownames = FALSE, options = list(dom = "t"))
```


```{r}
chain_data <- tidy_data %>%
  group_by(lipid_name, condition) %>%
  summarise(total_value = sum(value)) %>%
  ungroup() %>%
  separate(col = lipid_name, into = c("lipid_name", "chain_length"), sep = " ") %>%
  separate(col = chain_length, into = c("chain_length", "double_bonds"), sep = ":", convert = TRUE)

chain_data_all <- tidy_data %>%
  separate(col = lipid_name, into = c("lipid_name", "chain_length"), sep = " ") %>%
  separate(col = chain_length, into = c("chain_length", "double_bonds"), sep = ":", convert = TRUE) %>%
  group_by(condition, chain_length, double_bonds) %>%
  summarise(total_value = sum(value)) %>%
  ungroup()
  
chain_data_all_norm <- chain_data_all %>%
  left_join(condition_totals) %>%
  mutate(scaled_value = total_value * scale_factor)

chain_data_norm <- chain_data %>%
  left_join(condition_totals) %>%
  mutate(scaled_value = total_value * scale_factor) %>%
  mutate(log2_scaled = log2(scaled_value))
```

## All classes grouped together, scaled
```{r, fig.width = 10, fig.height=4}
chain_data_all_norm %>%
  drop_na() %>%
  ggplot(aes(x = chain_length, y = double_bonds, col = scaled_value, size = scaled_value)) +
  geom_point(position= position_dodge(width = 0.5), aes()) +
  scale_colour_viridis_b(direction = -1) +
  facet_wrap(~condition, scales = "free") 
```


## All classes grouped together, unscaled

```{r, fig.width = 10, fig.height=4}
chain_data_all %>%
  drop_na() %>%
  ggplot(aes(x = chain_length, y = double_bonds, col = total_value, size = total_value)) +
  geom_point(position= position_dodge(width = 0.5), aes()) +
  scale_colour_viridis_b(direction = -1) +
  facet_wrap(~condition, scales = "free") 
```


### TG - scaled values between conditions, linear 

```{r, fig.width = 10, fig.height=4}
chain_data_norm %>%
  drop_na() %>%
  filter(lipid_name == "TG") %>%
  #ggplot(aes(x = chain_length, y = double_bonds, col = log2_scaled, size = log2_scaled)) +
  ggplot(aes(x = chain_length, y = double_bonds, col = scaled_value, size = scaled_value)) +
  geom_point(position= position_dodge(width = 0.5), aes()) +
  scale_colour_viridis_b(direction = -1) +
  facet_wrap(~condition, scales = "free") 
```


```{r, fig.width = 10, fig.height=4, eval = FALSE}
# ### TG - scaled values between conditions, log2
chain_data_norm %>%
  drop_na() %>%
  filter(lipid_name == "TG") %>%
  ggplot(aes(x = chain_length, y = double_bonds, col = log2_scaled, size = log2_scaled)) +
  geom_point(position= position_dodge(width = 0.5), aes()) +
  scale_colour_viridis_b(direction = -1) +
  facet_wrap(~condition, scales = "free") 
```

### TG - unscaled, linear 

```{r, fig.width = 10, fig.height=4}
chain_data %>%
  drop_na() %>%
  #mutate(log2_value = log2(total_value)) %>%
  filter(lipid_name == "TG") %>%
  ggplot(aes(x = chain_length, y = double_bonds, col = total_value, size = total_value)) +
  geom_point(position= position_dodge(width = 0.5), aes()) +
  scale_colour_viridis_b(direction = -1) +
  facet_wrap(~condition, scales = "free") 

```

## All classes - chain length and double bonds (scaled values)

```{r, fig.width = 12, fig.height=40}
chain_data_norm %>%
  drop_na() %>%
  ggplot(aes(x = chain_length, y = double_bonds, col = scaled_value, size = scaled_value)) +
  geom_point(position= position_dodge(width = 0.5), aes()) +
  scale_colour_viridis_b(direction = -1) +
  facet_grid(vars(lipid_name), vars(condition), scales = "free") 
```



```{r, fig.width = 12, fig.height=10, eval = FALSE}
## Double bonds
chain_data %>%
  ggplot(aes(x = double_bonds, y = log2_value, col = condition)) +
  geom_point(position= position_dodge(width = 0.5)) +
  scale_color_manual(values = c("#9ac4d6", "#07634b", "#a961e8")) +
  facet_wrap(~class, scales = "free") 
```

## Experimental info

<div class = "row">
<div class = "col-md-6">
```{r}
DT::datatable(meta_type, rownames = FALSE, options = list(pageLength = 15, dom = "tp"))
```
</div>
</div>
<br>
   
### QQ plots of all data  
  
0 values have been removed.  

```{r, out.width=c('50%', '50%'), fig.show='hold', fig.height = 6}
removed_0 <- tidy_data %>%
  filter(value > 0)

qqnorm(removed_0$value, main = "linear/raw")
qqline(removed_0$value, col = "red", lwd = 2)

qqnorm(removed_0$log2_value, main = "log2 transformed")
qqline(removed_0$log2_value, col = "red", lwd = 2)
```
